---
title: "02_Differential_analysis"
author: "Aurelien Dugourd"
date: "5/12/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### License Info

This program is free software: you can redistribute it and/or modify it under 
the terms of the GNU General Public License as published by the Free Software 
Foundation, either version 3 of the License, or (at your option) any later 
version.

This program is distributed in the hope that it will be useful, but WITHOUT ANY 
WARRANTY; without even the implied warranty of MERCHANTABILITY or FITNESS FOR 
A PARTICULAR PURPOSE. See the GNU General Public License for more details.

Please check http://www.gnu.org/licenses/.

## Introduction

This is the second part in a series of transcriptomics tutorials. In
01_normalisation, we created a normalised dataset and summarised the
experimental design from which the data was generated. In the current tutorial,
we present examples of differential analysis of omics data using the normalised
RNAseq dataset to demonstrate.

## Getting Started

We first load the required libraries. 

```{r, message=FALSE}
#Main libraries
library(readr)
library(limma)

#Support functions also requires
library(ggplot2)
library(reshape)
library(pheatmap)
library(gridExtra)
library(grid)
library(cowplot)
library(ggrepel)
library(hexbin)

source("support_functions.R")
```

### Import the normalised dataframe and experimental design

```{r}
#Data
count_df_vsn <- as.data.frame(read_csv("../data/count_df_vsn.csv"))
row.names(count_df_vsn) <- count_df_vsn[,1]
count_df_vsn <- count_df_vsn[,-1]
#Design
targets <- as.data.frame(read_csv("../support/targets.csv"))
```

### LIMMA differential analysis
Now let's run a simple differential analysis using a simple wrapper for such situation `runLimma`.

```{r}
#first check the conditions order
unique(targets$condition)

#we want to compare the KO condition with the WT condition so we build a
#comparison list
comparisons <- list("KOvsWT" = c(2,-1)) #each vector of the list represent the contrasts, here we subtract the first condition (-1) from the second one (2)

#now that the comparisons are defined, we can run limma
limmaRes <- runLimma(measurements = count_df_vsn, 
                     targets = targets, 
                     comparisons = comparisons)

#once limma has run, we extract the statistics dataframe to summarise the
#differential analysis
ttop_KOvsWT <- ttopFormatter(topTable(limmaRes[[1]], coef = 1, number = length(count_df_vsn[,1]), adjust.method = "fdr"))

#make a qqplot
null_model <- pnorm(rnorm(length(ttop_KOvsWT[,1])))
plot(sort(null_model), sort(ttop_KOvsWT$P.Value) ,xlim = c(1,0), ylim = c(1,0)) #not bad, not great, let's proceed
abline(coef = c(0,1))
```

The qqplot (observed p-value distribution plotted against a random baseline) is
meant to give us an idea of the signal strength when comparing the KO and WT
conditions. The more the black dotted line deviates from the diagonal toward the
upper part of the plot, the stronger the signal is. This plot is very similar to
a p-value histogram.
In this case, we can observe a decent signal, but there are many contexts where
it could be much stronger. This type of difference is typical for single
mutations compared to a wild type, which usually lead to weak overall
differences. Stronger differences are usually expected for treatment vs control
experiments, or when comparing tumor vs healthy tissues.

### Write the DA output
```{r}
write_csv(ttop_KOvsWT, "../results/ttop_KOvsWT.csv")
```

## Session Info Details

```{r, echo=FALSE, eval=TRUE}
sessionInfo()
```






