---
title: "03_Pathway_activity_with_Progeny"
author: "Alberto Valdeolivas"
date: "13/05/2020"
output: github_document
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### License Info

This program is free software: you can redistribute it and/or modify it under 
the terms of the GNU General Public License as published by the Free Software 
Foundation, either version 3 of the License, or (at your option) any later 
version.

This program is distributed in the hope that it will be useful, but WITHOUT ANY 
WARRANTY; without even the implied warranty of MERCHANTABILITY or FITNESS FOR 
A PARTICULAR PURPOSE. See the GNU General Public License for more details.

Please check http://www.gnu.org/licenses/.

## Introduction

This is the third part in a series of transcriptomics tutorials. We present here how to estimate pathway activity from transcriptomics 
data using **PROGENy**. 

Conventional pathway analysis methods rely on the gene expression of the pathway 
members. However, this approach overlooks the effect of post-translational 
modifications and only captures very specific experimental conditions. 
To overcome these limitations, **PROGENy** (Pathway RespOnsive GENes) estimates 
the activity of relevant signaling pathways based on consensus gene signatures 
obtained from perturbation experiments [(Schubert et al. 2018)](https://www.nature.com/articles/s41467-017-02391-6)

**PROGENy** initially contained 11 pathways and was developed for the 
application to human transcriptomics data. It has been recently shown that 
**PROGENy** is also applicable to mouse data and it has been expanded to 
14 pathways [(Holland et al., 2019)](https://doi.org/10.1016/j.bbagrm.2019.194431). 
In addition, **PROGENy** can be applied to scRNA-seq data, as described in 
[(Holland et al., 2020)](https://doi.org/10.1186/s13059-020-1949-z)

**PROGENy** is available as a [Bioconductor package](http://bioconductor.org/packages/release/bioc/html/progeny.html). 
For additional information about the **PROGENy** method, visit its website: 

https://saezlab.github.io/progeny/

## Getting Started

We first load the required libraries. 

```{r libraries, message=FALSE}
library(progeny)
library(dorothea)
library(tibble)
library(tidyr)
library(dplyr)
library(ggplot2)
library(pheatmap)
library(readr)
```

In addition, we read the results from the previous scripts: 

+ Normalised transcriptomics data from (01_normalisation.Rmd)
+ Differential expression analysis results (02_differential_analysis.Rmd). 
In this particular case, a top table results object from limma. 

```{r loadInput, message=FALSE}
## We read the normalised counts and the experimental design 
Normalised_counts <- read_csv("../data/count_df_vsn.csv")
Experimental_design <- read_csv("../support/targets.csv")

## We read the results from the differential analysis. 
ttop_KOvsWT <- read_csv("../results/ttop_KOvsWT.csv")
```

We have to slightly modify the format of the input files to make it suitable
for running Progeny. 

```{r}
Normalised_counts_matrix <- Normalised_counts %>% 
    dplyr::mutate_if(~ any(is.na(.x)),~ if_else(is.na(.x),0,.x)) %>% 
    tibble::column_to_rownames(var = "gene") %>% 
    as.matrix()

ttop_KOvsWT_matrix <- ttop_KOvsWT %>% 
    dplyr::select(ID, t) %>% 
    dplyr::filter(!is.na(t)) %>% 
    column_to_rownames(var = "ID") %>%
    as.matrix()
```


## Pathway activity with Progeny

We first compute **Progeny** scores for every sample (with the replicates) 
using the normalised counts. It is worth noting that we are going to use the 100
most responsive genes per pathway. This number can be increased depending on 
the coverage of your experiments. For instance, the number of quantified genes
for single-cell RNA-seq is smaller than for Bulk RNA-seq or microarray. In those
cases, we suggest to increase the number of responsive genes to 200-500. 

```{r ProgenyCounts, message=FALSE}
PathwayActivity_counts <- progeny(Normalised_counts_matrix, scale=TRUE, 
    organism="Human", top = 100)
Activity_counts <- as.vector(PathwayActivity_counts)
```

We present the results in a heatmap:

```{r HeatmapProgeny, dpi=300}
paletteLength <- 100
myColor <- 
    colorRampPalette(c("darkblue", "whitesmoke","indianred"))(paletteLength)

progenyBreaks <- c(seq(min(Activity_counts), 0, 
    length.out=ceiling(paletteLength/2) + 1),
    seq(max(Activity_counts)/paletteLength, 
    max(Activity_counts), 
    length.out=floor(paletteLength/2)))

progeny_hmap <- pheatmap(t(PathwayActivity_counts),fontsize=14, 
    fontsize_row = 10, fontsize_col = 10, 
    color=myColor, breaks = progenyBreaks, 
    main = "PROGENy (100)", angle_col = 45,
    treeheight_col = 0,  border_color = NA)
```


Now, we run an enrichment analysis using a competitive permutation approach to 
assess the significance of the pathway activity. We end up with Normalised Enrichment Scores (NES) for each pathway.

```{r ProgenyStat, message=FALSE}
PathwayActivity_zscore <- progeny(ttop_KOvsWT_matrix, 
    scale=TRUE, organism="Human", top = 100, perm = 10000, z_scores = TRUE) %>%
    t()
colnames(PathwayActivity_zscore) <- "NES"
```

```{r BarplotProgeny, dpi=300}
PathwayActivity_zscore_df <- as.data.frame(PathwayActivity_zscore) %>% 
    rownames_to_column(var = "Pathway") %>%
    dplyr::arrange(NES) %>%
    dplyr::mutate(Pathway = factor(Pathway))

ggplot(PathwayActivity_zscore_df,aes(x = reorder(Pathway, NES), y = NES)) + 
    geom_bar(aes(fill = NES), stat = "identity") +
    scale_fill_gradient2(low = "darkblue", high = "indianred", 
        mid = "whitesmoke", midpoint = 0) + 
    theme_minimal() +
    theme(axis.title = element_text(face = "bold", size = 12),
        axis.text.x = 
            element_text(angle = 45, hjust = 1, size =10, face= "bold"),
        axis.text.y = element_text(size =10, face= "bold"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
    xlab("Pathways")
```

The MAPK pathway is the most active pathway upon the perturbation that we 
are studying (KO of the FOXA2 gene versus the wild-type). We can therefore 
visualise the MAPK most responsive genes (progeny weights) along with their 
t_values to interpret the results. In the scatterplot, we can see the genes
that are contributing the most to the pathway enrichment. 

```{r ProgenySccater_1, message=FALSE, warning=FALSE}
prog_matrix <- getModel("Human", top=100) %>% 
    as.data.frame()  %>%
    tibble::rownames_to_column("GeneID")

ttop_KOvsWT_df <- ttop_KOvsWT_matrix %>% 
    as.data.frame() %>% 
    tibble::rownames_to_column("GeneID")

scat_plots <- progeny::progenyScatter(df = ttop_KOvsWT_df, 
    weight_matrix = prog_matrix, 
    statName = "t_values", verbose = FALSE)
```

```{r ProgenySccater_2, dpi=300}
plot(scat_plots[[1]]$`MAPK`) 
```

Progeny results can be used as an optional input for **CARNIVAL**. CARNIVAL
sets weights based on **PROGENy** scores in each pathway-related node in order
to find more relevant solutions. We therefore run **PROGENy** again with 
slightly different parameters, setting `z_scores = FALSE` so that **PROGENy** returns pathway activity values between 1 and -1, rather than converting to Z-Scores.

```{r ProgenyCARNIVAL}
PathwayActivity_CARNIVALinput <- progeny(ttop_KOvsWT_matrix, 
    scale=TRUE, organism="Human", top = 100, perm = 10000, z_scores = FALSE) %>%
    t () %>% 
    as.data.frame() %>% 
    tibble::rownames_to_column(var = "Pathway") 
colnames(PathwayActivity_CARNIVALinput)[2] <- "score"
write_csv(PathwayActivity_CARNIVALinput, 
    "../results/PathwayActivity_CARNIVALinput.csv")
```


## References

> Schubert M, Klinger B, Klünemann M, Sieber A, Uhlitz F, Sauer S, Garnett MJ, Blüthgen N, Saez-Rodriguez J. “Perturbation-response genes reveal signaling footprints in cancer gene expression.” _Nature Communications_: [10.1038/s41467-017-02391-6](https://doi.org/10.1038/s41467-017-02391-6)

> Holland CH, Szalai B, Saez-Rodriguez J. "Transfer of regulatory knowledge from human to mouse for functional genomics analysis." _Biochimica et Biophysica Acta (BBA) - Gene Regulatory Mechanisms._ 2019. DOI: [10.1016/j.bbagrm.2019.194431](https://doi.org/10.1016/j.bbagrm.2019.194431).

> Holland CH, Tanevski J, Perales-Patón J, Gleixner J, Kumar MP, Mereu E, Joughin BA, Stegle O, Lauffenburger DA, Heyn H, Szalai B, Saez-Rodriguez, J. "Robustness and applicability of transcription factor and pathway analysis tools on single-cell RNA-seq data." _Genome Biology._ 2020. DOI: [10.1186/s13059-020-1949-z](https://doi.org/10.1186/s13059-020-1949-z).

## Session Info Details

```{r sessionInfo, echo=FALSE, eval=TRUE}
sessionInfo()
```
