---
title: "06_analysis_CARNIVAL_results"
author: "Rosa Hernansaiz-Ballesteros"
date: "18/07/2020"
output: 
  github_document: default
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### License Info

This program is free software: you can redistribute it and/or modify it under 
the terms of the GNU General Public License as published by the Free Software 
Foundation, either version 3 of the License, or (at your option) any later 
version.

This program is distributed in the hope that it will be useful, but WITHOUT ANY 
WARRANTY; without even the implied warranty of MERCHANTABILITY or FITNESS FOR 
A PARTICULAR PURPOSE. See the GNU General Public License for more details.

Please check http://www.gnu.org/licenses/.

## CARNIVAL output

CARNIVAL (CAusal Reasoning for Network identification using Integer VALue programming) 
returns a list with 4 elements: 

* weightedSIF. Character matrix. Contains a summary of _sifAll_, 
where *Weight* indicates the number of networks that present that interaction.

* nodesAttributes. Character matrix. Contains a summary of _attributesAll_, 
where all columns ending in *Act* indicate the number of networks that present that node in
each of the states (Zero = doesn't appear in the network, Up = upregulated, Down = downregulated, Avg = average regulation).
The column *NodeType* indicates if the node is a TF's (T) or a perturbed one (S).

* sifAll. List of character matrices. All networks that have been calculated.

* attributesAll. List of character matrices.

NOTE: This is valid for current CARNIVAL version (1.0.0)

## CARNIVAL interpretation

The summary files ( _sifAll_ and _attributesAll_ ) can be directly used to identify causal interactions between the 
perturbed nodes and the selected Transcription Factors. 
In addition to extract direct information from the network, 
we can run different downstream analysis based on the necesities of each project.

Here are described some of the downstream approaches that we have used:

+ Pathway enrichment analysis
+ Network comparisons

# Get ready

```{r libraries, message=FALSE}
library(readr)
library(piano)
library(dplyr)
library(ggplot2)
library(tibble)
library(tidyr)
library(dplyr)
library(ggplot2)
library(plyr)
library(GSEABase)
library(network)
library(reshape2)

# set working directory
setwd("/Users/rosherbal/Projects/transcriptutorial/scripts")

## We also load the support functions
source("support_enrichment.r")
source("support_networks.r")

## and the data

#read CARNIVAL results
carnival_result = readRDS("../results/carnival_result.rds")
carnival_sample_resolution = readRDS("../results/carnival_sample_resolution.rds")

```

# Enrichment Analysis

We define two different gene sets in order tor conduct the enrichment.
The first set contains the nodes that appear in the CARNIVAL output,
and are therefore relevant in the context of our input transcriptomic data. 
The second set contains all the genes in our prior knowledge network which are used as the backgroud.

We also downloaded from MSigDB https://www.gsea-msigdb.org/ the following dataset: c2.cp.v7.1.symbols.gmt. 
It contains several pathways from different resources and the genes that are known to be involved in those pathways.

Using the *Piano* R package, 
we run a gene set analysis (GSA) based on a list of significant genes (CARNIVAL nodes) and a gene set collection (background). 
It uses Fisher’s exact test.

```{r enrichment, message=FALSE}

# Load pathways
pathways = gmt_to_csv("../../Hepatoblastoma-iPC/proteomics/data/c2.cp.v7.1.symbols.gmt")

# Extract nodes and background
nodes_carnival = extractCARNIVALnodes(carnival_result)

# Run GSA hyper Geometric test
sig_pathways <- runGSAhyper(genes = nodes_carnival$sucesses, 
    universe = nodes_carnival$bg, gsc = loadGSC(pathways))
sig_pathways_df <- as.data.frame(sig_pathways$resTab)  %>% 
  tibble::rownames_to_column(var = "pathway") 

#data for plotting
PathwaysSelect <- sig_pathways_df %>%
    dplyr::select(pathway, `p-value`, `Adjusted p-value`) %>%
    dplyr::filter(`Adjusted p-value` <= 0.001) %>%
    dplyr::rename(pvalue = `p-value`, AdjPvalu = `Adjusted p-value`) %>% 
    dplyr::mutate(pathway = as.factor(pathway))

# Visualize top results
ggplot(PathwaysSelect, aes(x = reorder(pathway, pvalue), 
            y = -log10(pvalue))) + 
        geom_bar(stat = "identity") +
        scale_fill_gradient2(low = "darkblue", high = "indianred", 
            mid = "whitesmoke", midpoint = 0) + 
        coord_flip() +
        theme_minimal() +
        theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) + 
        xlab("")

```

# Network comparison

When we have more than 1 network, we usually would like to know how (dis)similar these networks are.
We can get a sense of the size of the reconstructed networks using topological parameters, such as number of edges and nodes.
We can also have a look at the network's density and the *degree distribution*.
The *density* indicates the proportion of interactions that exisit in our network when comparing with all posible interactions that can be stablished. 
The *degree distribution* shows the number of connections of a node. 
In a direct network, we can distinguish between incoming and outgoing connections.

NOTE: Here the density is calculated for a _direct graph_.
As CARNIVAL can repot 2 interactions between the same 2 nodes with different sign, 
these "doubled" interactions are excluded when calculating the density.


```{r netcompare, message=FALSE}

# get only summary files from CARNIVAL results
sifts = lapply(carnival_sample_resolution, function(x){x$weightedSIF})
nodos = lapply(carnival_sample_resolution, function(x){x$nodesAttributes})

# Calculate the number of edges and nodes in the networks and it's density
node_edge = do.call(rbind,lapply(sifts, count_edges_nodes_degree))

# Calculate degree distribution for a sample
count_degree = sifts$PANC1.FOXA2KO.Rep3 %>% degree_count

#degree distribution
p = data.frame(table(count_degree$total_count) / nrow(count_degree))
colnames(p) = c("Var1", "total_degree")
p = merge.data.frame(p, data.frame(table(count_degree$in_count) / nrow(count_degree)), all = T)
colnames(p) = c("Var1", "total_degree", "in_degree")
p = merge.data.frame(p, data.frame(table(count_degree$out_count) / nrow(count_degree)), all = T)
colnames(p) = c("k", "total_degree", "in_degree", "out_degree")
p = melt(p)
p$k = relevel(p$k, "0")
#visualize
ggdat = as.data.frame(node_edge) %>% tibble::rownames_to_column(var = "sample") %>%
  dplyr::mutate(condition = gsub(".Rep[0-9]{1}", "", sample))

#Ploting

# relation between number of edges and nodes
ggplot(ggdat, aes(x = nodes, y = edges, color = as.factor(condition))) +
  geom_point() +
  geom_text(label = ggdat$sample, 
            check_overlap = TRUE,
            vjust = 0, 
            nudge_y = 0.5,
            show.legend = F) +
  theme_bw(base_size = 15) +
  guides(color = guide_legend( title="Conditions" ) ) +
  ggtitle("Node-edge composition")

#networ degree
ggplot(ggdat, aes(x = density, y = sample, fill = as.factor(condition) )) +
  geom_col() +
  theme_bw(base_size = 15) +
  guides(fill = guide_legend( title="Conditions" ) ) +
  ggtitle("Network degree")

# degree distribution

ggplot(data = p, aes(x = as.factor(k), y = value, group= variable, color = variable)) +
  geom_point() +
  geom_line()  +
  theme_bw(base_size = 15) +
  guides(color = guide_legend( title="degree type" ) ) +
  ggtitle("Degree distribution")

```

## Session Info Details

```{r sessionInfo, echo=FALSE, eval=TRUE}
sessionInfo()
```
